Unprepared colonoscopy identifies microbiota specific to the proximal and distal human colon

### 2017-07-28 submission/manuscript.tex
v2 done for Pat

The human colon contains chemicals and nutrients that change along the
proximal to distal gut axis. These gradients create microenvironments
that affect the distrubution and composition of the gut microbiota. The
microbiome has been implicated in the colonic diseases colorectal cancer
(CRC) and inflammatory bowel disease (IBD). Further, these diseases
exhibit different symptoms depending on the location of the colon they
are found in. CRC tumors of the proximal and distal colon are
morphologically and genetically distinct. Similarly, bowel diseases such
as Crohn's are typically exacerbated in the proximal instestine while
ulcerative colitis patients often experience symptoms in the distal
colon. Previous analysis of the fecal microbiota from healthy and CRC or
IBD patients has revealed different microbial signatures associated with
these diseases. We extended these observations of the fecal microbiome
to include analysis of the proximal and distal healthy human colon. We
used a two-colonoscope approach on subjects that had not undergone
standard bowel preparation procedure. This technique allowed us to
characterize the native proximal and distal luminal and mucosal
microbiome without prior chemical disruption. 16S rRNA gene sequencing
was performed on proximal and distal mucosal and luminal biopsies and
home-collected stool for 20 healthy individuals. Diversity analysis
revealed that each site contained a diverse community, and that a
patient's samples were more similar to each other than to that of other
individuals. Comparison of all samples to fecal samples taken at exit
uncovered that the feces were most similar to samples taken from the
distal lumen, likely reflecting the anatomical structure of the colon.
Since we could not differentiate sites along the colon based on
community structure or community membership alone, we employed the
Random Forest machine-learning algorithm to identify key species that
distinguish biogeographical sites. Random Forest classification models
were built using taxa abundance and sample location and revealed
distinct populations that were found in each location.
\emph{Peptoniphilus, Anaerococcus, Enterobacteraceae, Pseudomonas} and
\emph{Actinomyces} were most likely to be found in mucosal samples
versus luminal samples (AUC = 0.925). The classification model performed
well (AUC = 0.912) when classifying mucosal samples into proximal or
distal sides, but separating luminal samples from each side proved more
challenging (AUC = 0.755). The distal mucosa was found to have high
populations of \emph{Finegoldia, Murdochiella} and \emph{Porphyromonas}.
Proximal and distal luminal samples were comprised of many of the same
taxa, likely reflecting the fact that stool moves along the colon from
the proximal to distal end. By sampling the unprepped human colon, our
results have identified distinct bacterial populations native to the
proximal and distal sides. Further investigation of these bacteria may
elucidate if and how these groups contribute to different pathogenesis
processes on the respective sides of the colon.

@@

The human colon is an ecosystem comprised of several different
microenvironments inhabited by resident bacterial members of the
microbiome. Concentrations of oxygen, water and anti-microbial peptides
change along the gut axis and influence what populations of microbes
reside in each location. Microenvironments differ not only
longitutinally along the colon, but latitudinally from the epithelium to
mucosa to intestinal lumen, offering several sites for different
microbial communities to flourish. The identity of these specific
microbes and communities are important for understanding the etiology of
complex colon diseases such as Inflammatory Bowel Disease (IBD) and
Colorectal Cancer (CRC). IBD and CRC are known to be preceded or
accelerated by perturbations in gut microbes ((1), (2), (3)). The
severity, symptoms, morbidity and mortality of these diseases is known
to vary based upon the biogeographical location in which they occur. For
instance, CRC tumors that arise on the distal side of the colon are
infiltrating lesions that present with painful symptoms ((4)). In
contrast, 47\% of CRCs are caused by proximal-sided colon tumors that
are sessile and form along the wall of the colon, often remaining
asymptomatic until advanced carcinogenesis ((4)). The distal and
proximal sides of the colon differ in the amount of inflammation present
and the genomic instability of precancerous cells, respectively, in
addition to variation in the previously mentioned chemical gradients
((1), (5), (6)). In IBD patients, disease flares in the distal colon are
usually indicative of ulcerative colitis (UC), whereas Crohn's disease
(CD) patients typically experience disease in the small intestine, ileum
and proximal colon ((2)). UC presents as large and highly inflammed
mucosal ulcers, where as CD lesions are often smaller and have areas of
normal tissue distributed amongst the flare ((2)). Thus, given the
varied physiology of the proximal-distal axis of the colon and known
differences in disease patterns at these sites, symbiotic microbes and
their metabolites likely vary as well, and may influence the
heterogenous disease prognoses of IBD and CRC. Because CRC is a
long-term complication of IBD, the distribution of microbes is important
to understanding the pathophysiology of both diseases.

Several recent findings have shown that development and progression of
IBD or CRC can be attributed to specific molecular events as a result of
interactions between the gut microbiota and human host ((1), (3)). For
instance, comparison of the bacteria present on CRC tumors with those
found on nearby healthy tissue has identified specific species that are
tumor-associated ((7)). These species include the oral pathogens
\emph{Fusobacterium nucleatum} and \emph{Porphyromonas asacharolytica}.
Interestingly, these periodontal pathogens have been highly predictive
of whether a patient had CRC tumors or not in our human stool
classification studies ((8)). \emph{F. nucleatum} has also been found to
be elevated in the stool and biopsies of patients with IBD as compared
to healthy controls (Strauss2011). Furthermore, studies of \emph{F.
nucleatum} isolated from mucosal biopsies showed that more invasive
\emph{F. nucleatum} positively correlates with IBD disease level
(Strauss2011). Like many intestinal pathogens, the bacteria appear to
have a high-impact despite being lowly-abundant in the community ((2)).
The molecular capabilities of these rare taxa may contribute to the
colonic disease state. These studies often examined only shed human
stool or the small intestine, preventing fine-resolution analysis of
paired samples from the proximal and distal sides of the colon.
Similarly, comparisons of on- or off-tumor/lesion bacteria rarely have
matched tissue from the other side of the colon from the same patient,
limiting what conclusions can be drawn about the colonic microbiome
overall, let alone at that specific site. Due to these limitations, the
contribution of the gut microbiota to IBD and CRC disease location in
the colon is largely undefined. Characterizing these communities could
provide needed insight into disease etiology, including how the
disruption of the healthy community could promote the initiation or
proliferation of the distinct proximal and distal CRC tumors or IBD
flares.

The few existing profiles of the microbial biogeography of the gut have
been limited by sample collection methods. The majority of human gut
microbiome studies have been performed on whole shed stool or on samples
collected during colonoscopy procedures. While the latter method allows
investigators to acquire samples from inside the human colon, typically
this procedure is preceded by the use of bowel preparation methods such
as the consumption of laxatives to cleanse the bowel. Bowel preparation
is essential for detecting cancerous or precancerous lesions in the
colon, but complicates microbiome profiling as the chemicals strip the
bowel of contents and disrupt the mucosal layer ((9), (10)). As such,
what little information we do have about the biogeographical
distribution of the microbes in the proximal and distal colon is
confounded by the bowel preparation procedure.

@@

Lumenal and mucosal samples were collected from the proximal and distal
colon of 20 healthy humans that had not undergone bowel preparation
(Figure 1). Participants also collected stool at home one week prior to
the procedure. To characterize the bacterial communities present at
these sites, 16S rRNA gene sequencing was performed on extracted DNA
from each sample. Each site was primarily dominated by \emph{Firmicutes}
and \emph{Bacteriodetes} (Figure 2A), consistent with known variability
in human microbiome research ((11)). Likewise, samples had varying
levels of diversity at each site, irrespective of the individual (Figure
2B). For example, the proximal mucosa was more diverse than the distal
for some individuals while the opposite was true for others. Therefore
we could not identify a clear pattern of changes in microbial diversity
along the gut axis.

@@

Next, we calculated \({\theta}\)YC distances to examine how each sample
compared to the home-collected exit stool. Amidst variability between
patients, we did identify significantly smaller \({\theta}\)YC distance
between the distal lumenal sample and the exit stool (Figure 3B, P
\textless{} 0.05, Wilcoxon, BH adjustment). Furthermore, there was an
even larger difference in the comparisons of the distal mucosa to the
exit stool, indicating that the mucosa is different from the stool as
compared to lumen (P \textless{} 0.0005, Wilcoxon, BH adjustment). To
determine what factors may be driving the differences seen among the
samples, we compared thetaYC distances between samples from all subjects
(interpersonal) versus samples from within one subject (intrapersonal).
We found that samples from one individual were far more similar to each
other than to other study subjects (Figure 3C), consistent with previous
human microbiome studies that have sampled multiple sites of the human
colon (\textbf{???}, (12), (13)). Thus interpersonal variation between
subjects drives the differences between samples more than sample site or
location. Overall, the results comparing the structure of the
communities suggest that the contents of the distal lumen are most
representative of stool at exit, and the microbes remaining on the
mucosa are much different.

@@

To identify OTUs that were distinct at each biogeographical site, we
constructed several Random Forest models trained using OTU abundances.
We built the first model to classify the lumen versus mucosal samples
for the proximal and distal sides, independently (Figure 4A). The
constructed model used 14 features for the proximal and 6 for the
distal. The models performed well when classifying these samples (0.8
and 1.0, respectively). The OTUs that were most predictive of each site
are identified by their greatest mean decrease in accuracy when removed
from the model. For distinguishing the proximal lumen and mucosa, OTUs
from the \emph{Bacteriodes}, \emph{Actinomyces}, \emph{Psuedomonas} and
two OTUs from the \emph{Enterobacteraceae} genera were differentially
abundant (Figure 4B). The model classifying the distal lumen and mucosa
identified OTUs from \emph{Turicibacter}, \emph{Finegoldia},
\emph{Peptoniphilus} and two OTUs from the \emph{Anaerococcus} genera
that could distinguish lumen from mucosal samples (Figure 4C). These
results indicate that there are fine differences between the different
sites of the colon, and that these can be traced down to specific OTUs
on each side.

Next, we built a model to differentiate the proximal and distal lumenal
samples. The model performed best when distinguishing the proximal
versus distal mucosa (Figure 5A, AUC = 0.9) compared to the proximal
versus distal lumen (AUC = 0.8). The model that separated proximal and
distal mucosal samples used 6 features. OTUs that were differentially
abundant between the distal and proximal mucosa included members of the
\emph{Porphyromonas}, \emph{Murdochiella}, \emph{Finegoldia},
\emph{Anaerococcus} and \emph{Peptoniphilus} genera (Figure 5B). The
model used 44 features to separate proximal and distal lumenal samples.
Differentially abundant OTUs of the proximal and distal lumen included
three OTUs of the \emph{Bacteroides} genus, a \emph{Clostridium IV} OTU
and an \emph{Oscillibacter} OTU (Figure 5C). This analysis found that
some of the same OTUs that are distinct between the mucosa and lumen
also helped to differentiate between the two sides- such as
\emph{Anaerococcus} and \emph{Finegoldia}.

@@

Given that specific bacterial species have been associated with
colorectal cancer and IBD, we probed our sample set for these OTUs.
Among our 100 samples, the most frequent sequence associated with the
\emph{Fusobacterium} genus was OTU179, which aligns via BLASTn to
\emph{Fusobacterium nucleatum subsp animalis} (100\% over full length).
This is the only species of \emph{Fusobacterium} known to have oncogenic
properties and be found on the surfaces of colorectal cancer tumors.
((14)). There were 14 samples with the \emph{F. nucleatum subsp.
animalis} sequence. Four sequences were located in the proximal mucosa
and three in the distal mucosal samples (Figure 6A). The second most
frequent \emph{Fusobacterium} sequence was OTU472, which aligns with
99\% identity to \emph{F. varium}. In addition to \emph{F. nucleatum},
\emph{F. varium} has been associated with IBD ((15)). 4 study
participants harbored \emph{F. varium} and the samples were split evenly
between the proximal and distal mucosa (Figure 6B). OTU152 was similar
to the members of the \emph{Porphyromonas} genus and the most frequent
sequence in that OTU aligned to \emph{Porphyromonas asacharolytica}
(99\% over full length), another bacterium commonly detected and
isolated from colorectal tumors. OTU152 was only detected on the distal
mucosa, and in fact was one of the OTUs the classification model
identified as separating distal and proximal sides (Figure 6C). Among
the samples that were positive for \emph{P. asacharolytica}, the
relative abundances for this OTU ranged from 0.01\% - 16\%. Thus,
disease-associated OTUs could be found in our sample set of 20 healthy
individuals.

@@

Here we identified bacterial taxa that were specific to the lumen and
mucosa of the proximal and distal sides of the human colon from samples
collected during unprepared colonoscopy. We found that all locations
contained a range of phyla and a range of diversity, but that there was
a wide variability between subjects. Pairwise comparisons of each of the
sites revealed that the proximal mucosa and lumen were most similar to
each other. Further, comparison of colonoscopy-collected samples with
samples collected from stool at home showed that the distal lumen is
most similar to stool at exit. Random Forest models built on OTU
relative abundances from each sample identified microbes that are
particular to each location of the colon. Finally, we were able to
detect some bacterial OTUs associated with colonic disease in our
healthy patient cohort. Using unprepped colonoscopy and machine
learning, we have identified bacterial phyla specific to the healthy
proximal and distal human colon.

When examining the relative abundance of the different phyla at each
site, there was a wide amount of variation for each phyla with
communities primarily dominated by the \emph{Bacteriodes} and
\emph{Firmicutes}. This likely reflects not only the variability between
human subjects, casued by differences in age, gender, diet, but also
reports of biogeogrphical ``patchiness'' in the gut microbiome. Several
studies have noted that the bacteria recoverable from the same mucosal
sample location can be vastly different when the samples are taken just
1 cm away from each other (16). Similar patchiness is also observed in
lumenal contents and fecal samples themselves; there is observed
separation of different interacting microbes along the length of a stool
sample, for instance (17). That said, across our samples the mucosal
samples harbor more \emph{Proteobacteria}, consistent with previous
studies comparing mucosal swabs to lumenal content in humans (5). Hence,
the conclusions we can draw from phyla analysis are likely impacted by
patchiness between subjects.

@@

We detected \emph{F. nucleatum} and \emph{P. asacharolytica} in 8 and 5
of our subjects, respectively. These bacteria have been shown to be
predictive of colorectal cancer in humans (8) and have oncogenic
properties in cell culture and in mice (18). Interestingly, while
\emph{F. nucleatum} was found on both sides of the colon, \emph{P.
asacharolytica} was only detected in the distal mucosa. Not much is
known about the distribution of \emph{P. asacharolytica} but given its
documented anaerobic characteristics and asacharolytic metabolism, it
might not be surprising that it resides in the less-oygen-rich and
proteinaceous distal mucosa (5). In studies examining bacteria on
colorectal cancer tumors, \emph{F. nucleatum} is more commonly detected
on proximal-sided tumors, and distribution of \emph{F. nucleatum}
decreases along the colon to rectum (19). Of the 8 (40\%) individuals
positive for \emph{F. nucleatum} in this study, the bacterium was spread
across the proximal mucosa, distal lumen and distal mucosa. Data
examining bacterial biofilms on the mucosa of CRC tumors suggests that
\emph{Fusobacteria} species are more commonly found on proximal tumors
and in biofilms, indicating that it is not only the presence of the
bacteria but the organization of the tumor community that contributes to
\emph{Fusobacterium's} role in tumorigenesis (7). Finally,
\emph{Fusobacterium} and \emph{Porphyromonas} species have been known to
not only co-occur on CRC tumors but also to synergistically promote
tumorigenesis in an oral cancer model (20), (21). Thus, further analysis
of the distribution and activities of these pathogens may elucidate a
mechanism for development of IBD or CRC subtypes in the proximal or
distal colon.

The \emph{Fusobacterium} species \emph{nucleatum} and \emph{varium} have
been commonly isolated from mucosal biopsies of patients with IBD (15).
Laboratory experiments with these isolates have shown that
disease-isolated \emph{F. nucleatum} are more invasive and stimulate
more TNF-a production than strains from healthy individuals (22),
suggesting the bacteria may increase inflammation in the gut as well
(23). \emph{F. varium} isolated from UC patients caused colonic ulcers
in an experimental mouse model (24). \emph{F. varium} was only detected
in 3 of our study participants and two of those samples were isolated
form the proximal mucosa (Figure 6B). Although there is low occurrance
in our study, \emph{F. varium} is most commonly isolated from UC patient
biopsies from the ileum or cecum (25), suggesting this species may
exhibit preference for the different environmental conditions of these
gastrointestinal sites. Further work will assess how gut environment may
select for species which then cause localized disease.

Specific comparisons of our findings to previously published gut
biogeography studies are additionally confounded by the use of bowel
preparation methods in most other studies. A rare report of a
matched-colonoscopy study sampled 18 patient's colonic mucosa and
lumenal contents prior to and after bowel cleansing (26). This group
found that mucosal and lumenal samples were distinguishable prior to
bowel cleansing, but that bowel preparation resulted in an increase in
shared OTUs between each site (26). Bowel cleansing not only made the
samples harder to distinguish, it resulted in decreases in diversity
across sites. Further, the differences were not great enough to overcome
interpersonal differences between subjects. Bowel preparation clearly
induces bias into the microbes recovered from sampling the lumen or
mucosa of a prepared bowel. Thus our findings in this study are
strengthened by the lack of bowel preparation.

@@

DNA extraction was performed using the PowerMicrobiome DNA/RNA Isolation
Kit (MO BIO Laboratories). For tissue biopsies, Bond-Breaker TCEP
solution (Fisher) and 2.8mm ceramic beads (MO BIO Laboratories) were
added to the bead beating step to enhance DNA recovery from mucosal
samples. The resulting DNA was used as template for amplification of the
V4 region of the 16S rRNA gene and fragments were sequenced on an
Illumina MiSeq as previously described (27). Sequences were curated
using the mothur software as described previously (28). The sequences
were assigned taxonomic classification using a naive Bayesian classifier
trained using a 16S rRNA gene training set from the Ribosomal Database
Project (RDP) (29) and clustered into operational taxonomic units (OTUs)
based on a 97\% similarity cutoff. Sequencing and analysis of a mock
community revealed the error rate to be 1.8\%. Samples were rarefied to
4231 sequences per sample in order to reduce uneven sampling bias.

>>>>

The human colon contains chemicals and nutrients that change along the
proximal to distal gut axis. These gradients create microenvironments
that affect the distrubution and composition of the gut microbiota. The
microbiome has been implicated in the colonic diseases colorectal cancer
(CRC) and inflammatory bowel disease (IBD). Further, these diseases
exhibit different symptoms depending on the location of the colon they
are found in. CRC tumors of the proximal and distal colon are
morphologically and genetically distinct. Similarly, inflammatory bowel
diseases such as Crohn's are typically exacerbated in the proximal
instestine while ulcerative colitis patients often experience symptoms
in the distal colon. Previous analysis of the fecal microbiota from
healthy and CRC or IBD patients has revealed different microbial
signatures associated with these diseases. We extended these
observations of the fecal microbiome to include analysis of the proximal
and distal healthy human colon. We used a two-colonoscope approach on
subjects that had not undergone standard bowel preparation procedure.
This technique allowed us to characterize the native proximal and distal
luminal and mucosal microbiome without prior chemical disruption. 16S
rRNA gene sequencing was performed on proximal and distal mucosal and
luminal biopsies and home-collected stool for 20 healthy individuals.
Diversity analysis revealed that each site contained a diverse community
and that a patient's samples were more similar to each other than to
that of other individuals. Comparison of all samples to fecal samples
taken at exit uncovered that the feces were most similar to samples
taken from the distal lumen, likely reflecting the anatomical structure
of the colon. Since we could not differentiate sites along the colon
based on community structure or community membership alone, we employed
the Random Forest machine-learning algorithm to identify key species
that distinguish biogeographical sites. Random Forest classification
models were built using taxa abundance and sample location and revealed
distinct populations that were found in each location. A model
differentiating the proximal mucosa and lumen was built with an AUC of
0.831. The proximal mucosa had a higher abundance of the genera
\emph{Enterobacteriaceae} and \emph{Bacteriodes}. \emph{Peptoniphilus,
Anaerococcus, Finegoldia,} and \emph{Turicibacter} were most likely to
be found in distal mucosal samples versus distal luminal samples (AUC =
0.98). The classification model performed well (AUC = 0.912) when
classifying mucosal samples into proximal or distal sides, but
separating luminal samples from each side proved more challenging (AUC =
0.755). The distal mucosa was found to have high populations of
\emph{Finegoldia, Murdochiella} and \emph{Porphyromonas}. Proximal and
distal luminal samples were comprised of many of the same taxa, likely
reflecting the fact that stool moves along the colon from the proximal
to distal end. By sampling the unprepped human colon, our results have
identified distinct bacterial populations native to the proximal and
distal sides. Further investigation of these bacteria may elucidate if
and how these groups contribute to different pathogenesis processes on
the respective sides of the colon.

@@

The human colon is an ecosystem comprised of several different
microenvironments inhabited by resident bacterial members of the
microbiome. Concentrations of oxygen, water and anti-microbial peptides
change along the gut axis and influence what populations of microbes
reside in each location. Microenvironments differ not only
longitutinally along the colon, but latitudinally from the epithelium to
mucosa to intestinal lumen, offering several sites for different
microbial communities to flourish. The identity of these specific
microbes and communities are important for understanding the etiology of
complex colon diseases such as Inflammatory Bowel Disease (IBD) and
Colorectal Cancer (CRC). IBD and CRC are known to be preceded or
accelerated by perturbations in gut microbes (1), (2), (3). The
severity, symptoms, morbidity and mortality of these diseases is known
to vary based upon the biogeographical location in which they occur. For
instance, CRC tumors that arise on the distal side of the colon are
infiltrating lesions that present with painful symptoms (4). In
contrast, 47\% of CRCs are caused by proximal-sided colon tumors that
are sessile and form along the wall of the colon, often remaining
asymptomatic until advanced carcinogenesis (4). The distal and proximal
sides of the colon differ in the amount of inflammation present and the
genomic instability of precancerous cells, respectively, in addition to
variation in the previously mentioned chemical gradients (1), (5), (6).
In IBD patients, disease flares in the distal colon are usually
indicative of ulcerative colitis (UC), whereas Crohn's disease (CD)
patients typically experience disease in the small intestine, ileum and
proximal colon (2). UC presents as large and highly inflammed mucosal
ulcers, where as CD lesions are often smaller and have areas of normal
tissue distributed amongst the flare (2). Thus, given the varied
physiology of the proximal-distal axis of the colon and known
differences in disease patterns at these sites, symbiotic microbes and
their metabolites likely vary as well, and may influence the
heterogenous disease prognoses of IBD and CRC. Because CRC is a
long-term complication of IBD, the distribution of microbes is important
to understanding the pathophysiology of both diseases.

Several recent findings have shown that development and progression of
IBD or CRC can be attributed to specific molecular events as a result of
interactions between the gut microbiota and human host (1), (3). For
instance, comparison of the bacteria present on CRC tumors with those
found on nearby healthy tissue has identified specific species that are
tumor-associated (7). These species include the oral pathogens
\emph{Fusobacterium nucleatum} and \emph{Porphyromonas asacharolytica}.
Interestingly, these periodontal pathogens have been highly predictive
of whether a patient had CRC tumors or not in our prior human stool
classification studies (8). \emph{F. nucleatum} has also been found to
be elevated in the stool and biopsies of patients with IBD as compared
to healthy controls (9). Furthermore, studies of \emph{F. nucleatum}
isolated from mucosal biopsies showed that more invasive \emph{F.
nucleatum} positively correlates with IBD disease level (9). Like many
intestinal pathogens, the bacteria appear to have a high-impact despite
being lowly-abundant in the community (2). The molecular capabilities of
these rare taxa may contribute to the colonic disease state. These
studies often examined only shed human stool or the small intestine,
preventing fine-resolution analysis of paired samples from the proximal
and distal sides of the colon. Similarly, comparisons of on- or
off-tumor/lesion bacteria rarely have matched tissue from the other side
of the colon from the same patient, limiting what conclusions can be
drawn about the colonic microbiome overall, let alone at that specific
site. Due to these limitations, the contribution of the gut microbiota
to IBD and CRC disease location in the colon is largely undefined.
Characterizing these communities could provide needed insight into
disease etiology, including how the disruption of the healthy community
could promote the initiation or proliferation of the distinct proximal
and distal CRC tumors or IBD flares.

The few existing profiles of the microbial biogeography of the gut have
been limited by sample collection methods. The majority of human gut
microbiome studies have been performed on whole shed stool or on samples
collected during colonoscopy procedures. While the latter method allows
investigators to acquire samples from inside the human colon, typically
this procedure is preceded by the use of bowel preparation methods such
as the consumption of laxatives to cleanse the bowel. Bowel preparation
is essential for detecting cancerous or precancerous lesions in the
colon, but complicates microbiome profiling as the chemicals strip the
bowel of contents and disrupt the mucosal layer (10), (11). As such,
what little information we do have about the biogeographical
distribution of the microbes in the proximal and distal colon is
confounded by the bowel preparation procedure.

@@

Lumenal and mucosal samples were collected from the proximal and distal
colon of 20 healthy humans that had not undergone bowel preparation
(Figure 1). Participants also collected stool at home one week prior to
the procedure. To characterize the bacterial communities present at
these sites, 16S rRNA gene sequencing was performed on extracted DNA
from each sample. Each site was primarily dominated by \emph{Firmicutes}
and \emph{Bacteriodetes} (Figure 2A), consistent with known variability
in human microbiome research (12). Likewise, samples had varying levels
of diversity at each site, irrespective of the individual (Figure 2B).
For example, the proximal mucosa was more diverse than the distal for
some individuals while the opposite was true for others. Therefore we
could not identify a clear pattern of changes in microbial diversity
along the gut axis.

@@

Next, we calculated \({\theta}\)YC distances to examine how each sample
compared to the home-collected exit stool. Amidst variability between
patients, we did identify significantly smaller \({\theta}\)YC distance
between the distal lumenal sample and the exit stool (Figure 3B, P
\textless{} 0.05, Wilcoxon, BH adjustment). Furthermore, there was an
even larger difference in the comparisons of the distal mucosa to the
exit stool, indicating that the mucosa is different from the stool as
compared to lumen (P \textless{} 0.0005, Wilcoxon, BH adjustment). To
determine what factors may be driving the differences seen among the
samples, we compared \({\theta}\)YC distances between samples from all
subjects (interpersonal) versus samples from within one subject
(intrapersonal). We found that samples from one individual were far more
similar to each other than to other study subjects (Figure 3C),
consistent with previous human microbiome studies that have sampled
multiple sites of the human colon (13), (14), (15). Thus interpersonal
variation between subjects drives the differences between samples more
than sample site or location. Overall, the results comparing the
structure of the communities suggest that the contents of the distal
lumen are most representative of stool at exit, and the microbes
remaining on the mucosa are much different.

@@

To identify OTUs that were distinct at each biogeographical site, we
constructed several Random Forest models trained using OTU abundances.
We built the first model to classify the lumen versus mucosal samples
for the proximal and distal sides, independently (Figure 4A). The
constructed model used 14 features for the proximal and 6 for the
distal. The models performed well when classifying these samples (0.831
and 0.98, respectively). The OTUs that were most predictive of each site
are identified by their greatest mean decrease in accuracy when removed
from the model. For distinguishing the proximal lumen and mucosa, OTUs
from the \emph{Bacteriodes}, \emph{Actinomyces}, \emph{Psuedomonas} and
two OTUs from the \emph{Enterobacteraceae} genera were differentially
abundant (Figure 4B). The model classifying the distal lumen and mucosa
identified OTUs from \emph{Turicibacter}, \emph{Finegoldia},
\emph{Peptoniphilus} and two OTUs from the \emph{Anaerococcus} genera
that could distinguish lumen from mucosal samples (Figure 4C). These
results indicate that there are fine differences between the different
sites of the colon, and that these can be traced to specific OTUs on
each side.

Next, we built a model to differentiate the proximal and distal lumenal
samples. The model performed best when distinguishing the proximal
versus distal mucosa (Figure 5A, AUC = 0.912) compared to the proximal
versus distal lumen (AUC = 0.755). The model that separated proximal and
distal mucosal samples used 6 features. OTUs that were differentially
abundant between the distal and proximal mucosa included members of the
\emph{Porphyromonas}, \emph{Murdochiella}, \emph{Finegoldia},
\emph{Anaerococcus} and \emph{Peptoniphilus} genera (Figure 5B). The
model used 44 features to separate proximal and distal lumenal samples.
Differentially abundant OTUs of the proximal and distal lumen included
three OTUs of the \emph{Bacteroides} genus, a \emph{Clostridium IV} OTU
and an \emph{Oscillibacter} OTU (Figure 5C). This analysis found that
some of the same OTUs that are distinct between the mucosa and lumen
also helped to differentiate between the two sides- such as
\emph{Anaerococcus} and \emph{Finegoldia}.

@@

Given that specific bacterial species have been associated with
colorectal cancer and IBD, we probed our sample set for these OTUs.
Among our 100 samples, the most frequent sequence associated with the
\emph{Fusobacterium} genus was OTU179, which aligns via BLASTn to
\emph{Fusobacterium nucleatum subsp animalis} (100\% over full length).
This is the only species of \emph{Fusobacterium} known to have oncogenic
properties and be found on the surfaces of colorectal cancer tumors.
(16). There were 14 samples with the \emph{F. nucleatum subsp. animalis}
sequence. Four sequences were located in the proximal mucosa and three
in the distal mucosal samples (Figure 6A). The second most frequent
\emph{Fusobacterium} sequence was OTU472, which aligns with 99\%
identity to \emph{F. varium}. In addition to \emph{F. nucleatum},
\emph{F. varium} has been associated with IBD (17). 4 study participants
harbored \emph{F. varium} and the samples were split evenly between the
proximal and distal mucosa (Figure 6B). OTU152 was similar to the
members of the \emph{Porphyromonas} genus and the most frequent sequence
in that OTU aligned to \emph{Porphyromonas asacharolytica} (99\% over
full length), another bacterium commonly detected and isolated from
colorectal tumors. OTU152 was only detected on the distal mucosa, and in
fact was one of the OTUs the classification model identified as
separating distal and proximal sides (Figure 6C). Among the samples that
were positive for \emph{P. asacharolytica}, the relative abundances for
this OTU ranged from 0.01\% - 16\%. Thus, disease-associated OTUs could
be found in our sample set of 20 healthy individuals.

@@

Here we identified bacterial taxa that were specific to the lumen and
mucosa of the proximal and distal sides of the human colon from samples
collected during unprepared colonoscopy. We found that all locations
contained a range of phyla abundances and a range of diversity, but that
there was a wide variability between subjects. Pairwise comparisons of
each of the sites revealed that the proximal mucosa and lumen were most
similar to each other. Further, comparison of colonoscopy-collected
samples with samples collected from stool at home showed that the distal
lumen is most similar to stool at exit. Random Forest models built on
OTU relative abundances from each sample identified microbes that are
particular to each location of the colon. Finally, we were able to
detect some bacterial OTUs associated with colonic disease in our
healthy patient cohort. Using unprepped colonoscopy and machine
learning, we have identified bacterial taxa specific to the healthy
proximal and distal human colon.

When examining the relative abundance of the different phyla at each
site, there was a wide amount of variation with communities primarily
dominated by the \emph{Bacteriodes} and \emph{Firmicutes}. This likely
reflects not only the variability between human subjects, casued by
differences in age, gender, diet, but also reports of biogeogrphical
``patchiness'' in the gut microbiome. Several studies have noted that
the bacteria recoverable from the same mucosal sample location can be
vastly different when the samples are taken just 1 cm away from each
other (18). Similar patchiness is also observed in lumenal contents and
fecal samples themselves; there is observed separation of different
interacting microbes along the length of a stool sample, for instance
(19). That said, across our samples the mucosal samples harbor more
\emph{Proteobacteria}, consistent with previous studies comparing
mucosal swabs to lumenal content in humans (5). Hence, the conclusions
we can draw from phyla analysis are likely impacted by patchiness
between subjects.

@@

We detected \emph{F. nucleatum} and \emph{P. asacharolytica} in 8 and 5
of our subjects, respectively. These bacteria have been shown to be
predictive of colorectal cancer in humans (8) and have oncogenic
properties in cell culture and in mice (20). Interestingly, while
\emph{F. nucleatum} was found on both sides of the colon, \emph{P.
asacharolytica} was only detected in the distal mucosa. Not much is
known about the distribution of \emph{P. asacharolytica} but given its
documented anaerobic characteristics and asacharolytic metabolism, it
might not be surprising that it resides in the less-oygen-rich and
proteinaceous distal mucosa (5). In studies examining bacteria on
colorectal cancer tumors, \emph{F. nucleatum} is more commonly detected
on proximal-sided tumors, and distribution of \emph{F. nucleatum}
decreases along the colon to rectum (21). Of the 8 (40\%) individuals
positive for \emph{F. nucleatum} in this study, the bacterium was spread
across the proximal mucosa, distal lumen and distal mucosa. Data
examining bacterial biofilms on the mucosa of CRC tumors suggests that
\emph{Fusobacteria} species are more commonly found on proximal tumors
and in biofilms, indicating that it is not only the presence of the
bacteria but the organization of the tumor community that contributes to
\emph{Fusobacterium's} role in tumorigenesis (7). Finally,
\emph{Fusobacterium} and \emph{Porphyromonas} species have been known to
not only co-occur on CRC tumors but also to synergistically promote
tumorigenesis in an oral cancer model (22), (23). Thus, further analysis
of the distribution and activities of these pathogens may elucidate a
mechanism for development of IBD or CRC subtypes in the proximal or
distal colon.

The \emph{Fusobacterium} species \emph{nucleatum} and \emph{varium} have
been commonly isolated from mucosal biopsies of patients with IBD (17).
Laboratory experiments with these isolates have shown that
disease-isolated \emph{F. nucleatum} are more invasive and stimulate
more TNF-a production than strains from healthy individuals (9),
suggesting the bacteria may increase inflammation in the gut as well
(24). \emph{F. varium} isolated from UC patients caused colonic ulcers
in an experimental mouse model (25). \emph{F. varium} was only detected
in three of our study participants and two of those samples were
isolated form the proximal mucosa (Figure 6B). Although there is low
occurrance in our study, \emph{F. varium} is most commonly isolated from
UC patient biopsies from the ileum or cecum (26), suggesting this
species may exhibit preference for the different environmental
conditions of these gastrointestinal sites. Further work will assess how
gut environment may select for species which may then cause localized
disease.

Specific comparisons of our findings to previously published gut
biogeography studies are additionally confounded by the use of bowel
preparation methods in most other studies. A rare report of a
matched-colonoscopy study sampled 18 patient's colonic mucosa and
lumenal contents prior to and after bowel cleansing (27). This group
found that mucosal and lumenal samples were distinguishable prior to
bowel cleansing, but that bowel preparation resulted in an increase in
shared OTUs between each site (27). Bowel cleansing not only made the
samples harder to distinguish, it resulted in decreases in diversity
across sites. Further, the differences were not great enough to overcome
interpersonal differences between subjects. Bowel preparation clearly
induces bias into the microbes recovered from sampling the lumen or
mucosa of a prepared bowel. Thus our findings in this study are
strengthened by the lack of bowel preparation.

@@

DNA extraction was performed using the PowerMicrobiome DNA/RNA Isolation
Kit (MO BIO Laboratories). For tissue biopsies, Bond-Breaker TCEP
solution (Fisher) and 2.8mm ceramic beads (MO BIO Laboratories) were
added to the bead beating step to enhance DNA recovery from mucosal
samples. The resulting DNA was used as template for amplification of the
V4 region of the 16S rRNA gene and fragments were sequenced on an
Illumina MiSeq as previously described (28). Sequences were curated
using the mothur software as described previously (29). The sequences
were assigned taxonomic classification using a naive Bayesian classifier
trained using a 16S rRNA gene training set from the Ribosomal Database
Project (RDP) (30) and clustered into operational taxonomic units (OTUs)
based on a 97\% similarity cutoff. Sequencing and analysis of a mock
community revealed the error rate to be 1.8\%. Samples were rarefied to
4231 sequences per sample in order to reduce uneven sampling bias.

Diversity analysis was performed using the Simpson diversity calculator
and thetaYC calculator metrics in mothur version 3.2 (29). ThetaYC
distances were calculated to determine the dissimilarity between two
samples. Random Forest classification models were built using the
randomForest R package and resultant models were used to identify the
OTUs that were most important for classifying each location (31). To get
species-level information about sequences of interest, sequences were
aligned using blastn and the species name was only used if the identity
score was \textgreater{}= 99\%.
